Accelerating structure–function mapping using the ViVa webtool to mine natural variation

### 2018-11-12 manuscript/01-intro.Rmd
clean up and rewrite of results section

The first genome sequence of *Arabidopsis thaliana* facilitated rapid advancement of plant biology through molecular genetics. Since this initial genome, massive scaling of sequencing technology has allowed a pioneering group to survey the genomic variation in natural *Arabidopsis thaliana* populations. This valuable population genetics resource has led to several associations of genetic loci with phenotypic traits and provided insights into how selective pressure has influenced the evolution of plant genomes. Outside of genome-wide studies however, this valuable resequencing dataset has seen little use.

Natural genomic variation also provides a set of feasible functional variation at the gene, gene family and gene network level. Characterizing this variation can provide insights into the function and evolution of genes, families and networks [@delker_natural_2010; @wright_insights_2017] . Here, we present a web application and R-package for gene, gene family and gene network analysis of this dataset. This application allows plant molecular biologists without bioinformatics experience to use genomic variation to formulate hypotheses as to the sequence/function/phenotype relationships within their gene, family, or network of their interest.

This application allows easy access to  
1. lists of missense polymorphisms to facilitate biochemical assays of variant effects [@starita_variant_2017],  
2. family-wise alignments of variants to facilitate de novo functional domain identification [@melamed_combining_2015],  
3. lists of accessions containing missense (or any type of) polymorphisms to facilitate segregation analysis and measurment of the distribution of variant effects on phenotype [@park_distributions_2017],  
4. as well as the study of gene and network evolution [@delker_natural_2010; @kliebenstein_role_2008].

Here, we demonstrate the utility of this tool through comprehensive analysis of genomic variation in the nuclear auxin signaling pathway. We identify potential functional variation within the coding sequence of the genes within this network and formulate hypotheses regarding the functional implications of this variation and the evolution of these gene families. Our analysis provides further confirmation of much of the existing knowledge of these genes demonstrating the validity of this approach. We also identify natural alleles of several genes and accessions bearing these alleles to facilitate future characterization structure/function, genotype/phenotype, and epistatic relationships.

>>>>

The first genome sequence of *Arabidopsis thaliana* facilitated rapid advancement of plant biology [@thearabidopsisgenomeinitiative_analysis_2000]. Since this initial genome, massive scaling of sequencing technology has allowed the survey the genomic variation in natural *Arabidopsis thaliana* populations [@nordborg_pattern_2005; @borevitz_genomewide_2007; @weigel_1001_2009]. This valuable population genetics resource has led to several associations of genetic loci with phenotypic traits and provided insights into how selective pressure has influenced the evolution of plant genomes [@long_massive_2013; @atwell_genomewide_2010; @clark_common_2007]. Despite the high value of this compendium of genetic variation, its use has been largely limited to investigations of population genetics and for genome-wide association studies.

Natural genetic variation provides a catalog of permissible polymorphisms which may facilitate exploration and prediction of genotype/function/phenotype relationships at the gene, gene family and network scales [@joly-lopez_developing_2016]. Outside of large effect variants, these relationships are difficult to map. Such studies in plants have the added obstacles of large gene families and the still quite time and resource-intensive process of allele replacement. Massively parallel assays of variant effects have revolutionized genetic diagnostics and personalized medicine [@starita_variant_2017; @gasperini_power_2016; @matreyek_platform_2017]. We envision the use of plant natural variation datasets as a tool to similarly revolutionize breeding and genetic engineering of crop plants by rapidly advancing our understanding of genotype/function/phenotype relationships. Biochemical or synthetic assays of gene function and quantitative phenotyping of the germplasm of collections of natural variation can facilitate functional and phenotypic quantification [@wright_insights_2017]. However, the high dimensionality of natural variation datasets often limits access to those with bioinformatics expertise.

Here, we present ViVa: a web application and R-package for Visualizing Variation, which allows plant molecular biologists of any level access to the 1001 genomes database. ViVa facilitates formulation of hypotheses as to the genotype/function/phenotype relationships within any gene, family, or network. Using ViVa researchers may: 1. Identify polymorphisms to facilitate biochemical assays of variant effects [@starita_variant_2017; @wright_insights_2017]; 2. Produce family-wise alignments of variants to facilitate de novo functional domain identification [@melamed_combining_2015]; 3. Generate lists of accessions containing polymorphisms to facilitate phenotypic analysis of gene variant effects [@park_distributions_2017]; and 4. Quantify metrics of genetic diversity to facilitate the study of gene, gene family and network evolution [@delker_natural_2010; @kliebenstein_role_2008].

### 2018-11-12 manuscript/03-methods.Rmd
clean up and rewrite of results section

### Variant Data
Variant data was queried from the 1001 genomes project website (http://1001genomes.org) via URL requests to an (undocumented) API service that provides subset files of the whole-genome VCF. The whole-genome VCF file can be found on the project’s website at http://1001genomes.org/data/GMI-MPI/releases/v3.1/

>>>>

### Variant Data
Variant data was queried from the 1001 genomes project (http://1001genomes.org) via URL requests to their API service (http://tools.1001genomes.org/api/index.html). These queries provided subsets of the whole-genome VCF. The whole-genome VCF file can be found on the project’s website at http://1001genomes.org/data/GMI-MPI/releases/v3.1/

### 2018-11-12 manuscript/Old sections/IAAs.Rmd
clean up and rewrite of results section

```{r IAAtree, fig.height=5, fig.width=5, fig.cap="(ref:IAAtree)"}
p <- ggtree(tree, layout = "rectangular", color = "grey") %<+% IAA_data + theme_tree() + scale_x_continuous(expand = c(0, 1.2)) + 
  geom_tippoint(mapping = aes(size = 4*PiN_PiS, fill = PiN_PiS), shape = 21, color = "white", show.legend = TRUE) + 
  scale_fill_viridis(values = rescale(c(min(IAA_data$PiN_PiS), 1,
                                        max(IAA_data$PiN_PiS)), 
                                      to = c(0,1)), guide = "colourbar") + scale_size(guide = "none") + 
  theme(legend.position = c(.9,.7)) + 
  geom_nodelab(mapping = aes(label = round(prob,digits = 2)), nudge_x = -.3, nudge_y = .5, size = 3) + 
  geom_tiplab(mapping = aes(color = Class), hjust = -0.25) + 
  scale_color_viridis(discrete = TRUE)
p
```

@@

(ref:A-IAA-alignment) 
```{r A-IAA-alignment, fig.cap="(ref:A-IAA-alignment)"}
A_IAAs <- subset(gene_info, tair_symbol %in% IAA_data[which(IAA_data$class == "A"), "name"])
A_ARFs <- alignCDS(A_IAAs$transcript_ID)
# make an alignment data frame
A_IAAs <- makeAlnDF(A_IAAs[[2]])
A_IAAs <- left_join(A_IAAs, dplyr::select(gene_info, "tair_locus",
                                       "tair_symbol", "transcript_ID"),
                        by = c("transcript_ID" = "transcript_ID"))
A_IAAs$tair_symbol %<>% as.factor()
A_IAAs$tair_symbol %<>% fct_reorder(
    {as.integer(str_remove(A_IAAs$tair_symbol, 
                           pattern = "ARF"))}, 
    .desc = TRUE)
A_IAAs <- addSNPsToAlnDF(A_IAAs, VCF)

p <-ggplot(A_IAAs, aes(x = aln_pos, y = tair_symbol,
                             group = seq_pos, text = variants)) +
  geom_raster(data = A_IAAs[A_IAAs$seq_pos == "-",], fill = "lightgrey") +
  geom_raster(data = na.omit(A_IAAs), 
             # mapping = aes(fill = effects)) +
              mapping = aes(fill = as.integer(as.factor(effects)))) +
  #scale_fill_viridis(option = "A", discrete = TRUE) +
   scale_fill_viridis(breaks = range(
     as.integer(as.factor(na.omit(A_IAAs$effects)))),
                      labels = c("deleterious", "neutral"),
                      option = "A", 
                      guide = guide_colorbar(ticks = FALSE, 
                                             title.vjust = .8)) +
  scale_y_discrete() +
  # expand increases distance from axis
  labs(x = "codon position in alignment", y = "", fill = "variant effect") +
  theme_logo(base_family = "Helvetica") +
  theme(panel.grid = element_blank(), panel.grid.minor = element_blank()) +
  theme(strip.background = element_blank(),
        strip.text.x = element_blank(),
        legend.position = "bottom")  
p
```

>>>>

```{r IAAtree, fig.height=5, fig.width=5, fig.cap="(ref:IAAtree)"}
p <- ggtree(tree, layout = "rectangular", color = "grey") %<+% IAA_data + theme_tree() + scale_x_continuous(expand = c(0, 1.2)) + 
  geom_tippoint(mapping = aes(size = 4*PiN_PiS, fill = PiN_PiS), shape = 21, color = "white", show.legend = TRUE) + 
  scale_fill_viridis(values = rescale(c(min(IAA_data$PiN_PiS), 1,
                                        max(IAA_data$PiN_PiS)), 
                                      to = c(0,1)), guide = "colourbar") + scale_size(guide = "none") + 
  theme(legend.position = c(.9,.7)) + 
  geom_nodelab(mapping = aes(label = round(as.numeric(prob),digits = 2)), nudge_x = -.3, nudge_y = .5, size = 3) + 
  geom_tiplab(mapping = aes(color = Class), hjust = -0.25) + 
  scale_color_viridis(discrete = TRUE)
p
```

@@

(ref:A-IAA-alignment) 
```{r A-IAA-alignment, fig.cap="(ref:A-IAA-alignment)"}
A_IAAs <- subset(gene_info, tair_symbol %in% IAA_data[which(IAA_data$Class == "A"), "name"])
A_IAAs <- alignCDS(A_IAAs$transcript_ID)
# make an alignment data frame
A_IAAs <- makeAlnDF(A_IAAs[[2]])
A_IAAs <- left_join(A_IAAs, dplyr::select(gene_info, "tair_locus",
                                       "tair_symbol", "transcript_ID"),
                        by = c("transcript_ID" = "transcript_ID"))
A_IAAs$tair_symbol %<>% as.factor()
A_IAAs$tair_symbol %<>% fct_reorder(
    {as.integer(str_remove(A_IAAs$tair_symbol, 
                           pattern = "IAA"))}, 
    .desc = TRUE)
VCF <- ldply(.data = VCF, .fun = subset, gt_GT != "0|0")
coding_VCF <- getCodingDiv(VCF)
A_IAAs <- addSNPsToAlnDF(aln_df = A_IAAs,SNPs = coding_VCF, effect_order = SNPeff_order)
A_IAAs$gap <- A_IAAs$seq_pos == "-"

p <-ggplot(A_IAAs, aes(x = aln_pos, y = tair_symbol)) +
  geom_raster(data = A_IAAs, 
             # mapping = aes(fill = effects)) +
              mapping = aes(fill = strength,
                            alpha = gap)) +
  scale_alpha_manual(values = c("TRUE" = 0, "FALSE" = 1), 
                     labels = c("aligned", "gap"), 
                     guide = guide_legend(override.aes = 
                                            list(fill = c("grey85", "white")),
                       title = "alignment",
                       direction = "vertical")) + 
   scale_fill_viridis(breaks = range(SNPeff_order$strength),
                      limits = range(SNPeff_order$strength),
                      na.value = "grey85",
                      direction = -1,
                      labels = c("neutral", "deleterious"),
                      option = "A", 
                      guide = guide_colorbar(ticks = FALSE, 
                                             title.vjust = .8, 
                                             title.position = "top", 
                                             title.hjust = 0.5)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_x_continuous(expand = c(0,0)) +  # expand increases distance from axis
  labs(x = "codon position in alignment", y = "", fill = "variant effect") +
  theme_logo(base_family = "Helvetica") +
  theme(panel.grid = element_blank(), 
        panel.grid.minor = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank(),
        #panel.background = element_rect(fill = "grey85", color = "white"),
        legend.position = c("bottom"),
        legend.key = element_rect(linetype = "solid"),
        legend.spacing = unit(4, "char"), 
        legend.box.spacing = unit(0.1,"line")) 
p

ggsave("A_IAAs.png", width = 7, height = 4)
```

### 2018-11-12 manuscript/Old sections/discussion.Rmd
clean up and rewrite of results section

Our analysis of the natural diversity in the gene families that make up the auxin nuclear signaling pathway has confirmed existing knowledge, supported existing hypothses and generated new hypotheses. These results have been generated by a team of undergraduate and graduate students, postdoctoral researchers, and principal investigators as an alpha-test of the web application developed here. Our analysis has confirmed the importance of auxin-induced degradation of the Aux/IAA proteins as evidenced by the strong conservation of the *AFB* genes and the canonical degron region of the *Aux/IAA* genes. Despite this relative conservation, this analysis has also identified many accessions containing new alleles of interest in the *AFB* and *Aux/IAA* families. 
-interesting points from the AFBs
-interesting points IAAs
-TPLs
-ARFs
  -DBD and PB1

>>>>

Our analysis of the natural diversity in the gene families that make up the auxin nuclear signaling pathway has confirmed existing knowledge, supported existing hypotheses and generated new hypotheses. These results have been generated by a team of undergraduate and graduate students, postdoctoral researchers, and principal investigators as an alpha-test of the ViVa web application developed here. Our analysis has confirmed the importance of auxin-induced degradation of the Aux/IAA proteins as evidenced by the strong conservation of the *AFB* genes and the canonical degron region of the *Aux/IAA* genes. Despite this relative conservation, this analysis has also identified many accessions containing new alleles of interest in the *AFB* and *Aux/IAA* families. 
-interesting points from the AFBs
-interesting points IAAs
-TPLs
-ARFs
  -DBD and PB1

### 2018-11-12 manuscript/index.Rmd
clean up and rewrite of results section

---
title: Predicting structure--function relationships using natural genetic variation 
author:  |
  | R. Clay Wright^[Corresponding author: wrightrc@vt.edu] $^1$, 
  | Morgan O. Hamm $^2$, 
  | Jennifer Nemhauser $^2$
  | $^1$Biological Systems Engineering, Virginia Tech, Blacksburg, VA 24061, 
  | $^2$Deparment of Biology, University of Washington, Seattle, WA 98195

@@

abstract: |
  The way genes and genomes have evolved can provide information about which genes are critical to an organisms fitness, which genes faciliate adaptation, and how robust genes and gene networks are to perturbation. Thousands of genomes have now been sequenced representing a large swath of plant species and genetic variation within many species which has only begun to be explored in the last few years. However, much of this data remains inaccessible to researchers without significant bioinformatics experience. Here, we have developed a webtool ViVa (Visualization of Variation) to facilitate analysis, by any researcher, of The 1001 *Arabidopsis* Genomes Project dataset on a gene, gene family, or gene network level. We We have demonstrated the utility and accessibility of this tool by analyzing the natural variation in the well-studied nuclear auxin signaling pathway as a group of researchers from an array of experience levels. Our analysis has provided further confirmation of existing knowledge of this well studied pathway and has also helped generate new hypotheses. These results suggest there is much to be learned about less studied gene families and networks from similar analysis, especially when paired with biochemical and genetic characterization. This application provides access for every researcher allowing the community to extract maximum value from this large and expensive-to-generate dataset. ViVa is also extensible to any other database of genetic variation such as The 3,000 Rice Genomes Project.

@@

#devtools::install_github("wrightrc/r1001genomes", ref = "auxin-natural-variation")
library(bookdown)
library(ape)
library(r1001genomes)
library(DT)
library(tidyverse)
library(magrittr)
library(RColorBrewer)
library(ggpmisc)
library(ggthemes)
library(ggseqlogo)
library(ggtree)
library(reticulate)
library(scales)
library(viridis)
library(ips)
library(knitr)
library(dplyr)

@@

## List of working titles  
* Predicting structure--function relationships using natural genetic variation 
* Accelerating structure--function mapping using natural genetic variation 
* A comprehensive analysis of natural sequence variation within the <i>Arabidopsis thaliana</i> nuclear auxin signaling pathway 
* **Add your suggestions**

>>>>

---
title: Accelerating structure–function mapping using the ViVa webtool to mine natural variation 
author:  |
  | R. Clay Wright^[Corresponding author: wrightrc@vt.edu] $^1$, 
  | Morgan O. Hamm $^2$,
  | Jennifer Nemhauser $^2$
  | $^1$Biological Systems Engineering, Virginia Tech, Blacksburg, VA 24061, 
  | $^2$Deparment of Biology, University of Washington, Seattle, WA 98195

@@

abstract: Thousands of sequenced genomes are now publically available and represent a large swath of plant evolution; yet, much of this data remains inaccessible to researchers without significant bioinformatics experience. Here, we present a webtool called ViVa (Visualization of Variation) which aims to empower any researcher to take advantage of the amazing genetic resource collected in the Arabidopsis 1001 Genomes Project. ViVa facilates data mining on the gene, gene family or gene network level. To test the utility and accessibility of ViVa, we assembled a team with a range of expertise with biology and bioinformatics to analyze the natural variation in the well-studied nuclear auxin signaling pathway. Our analysis has provided further confirmation of existing knowledge and has also helped generate new hypotheses regarding this well studied pathway. These results highlight how natural variation can be used to generate and test hypotheses about less studied gene families and networks, especially when paired with biochemical and genetic characterization. ViVa is also readily extensible to newly emerging databases of genetic variation, such as the 3,000 Rice Genomes Project.

@@

#devtools::install_github("wrightrc/r1001genomes", ref = "auxin-natural-variation")
library(bookdown)
library(ape)
library(r1001genomes)
library(DT)
library(tidyverse)
library(magrittr)
library(RColorBrewer)
library(ggpmisc)
library(gginnards)
library(ggthemes)
library(ggseqlogo)
library(ggtree)
library(reticulate)
library(scales)
library(viridis)
library(ips)
library(knitr)
library(dplyr)
library(stringr)

@@

## List of working titles  
* Predicting structure--function relationships using natural genetic variation 
* Accelerating structure--function mapping using natural genetic variation 
* A comprehensive analysis of natural sequence variation within the <i>Arabidopsis thaliana</i> nuclear auxin signaling pathway 
* Visualizing natural variation with ViVa, a new webtool
* Accelerating structure–function mapping using a new webtool to mine natural variation
